Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-02 response.tex

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove genes with low counts.  In practice, it is common to remove low count genes from RNA-seq datasets prior to analysis because these genes contain little information about differential expression and lead to computational difficulties for some methods. With that being said, we did not remove any genes from the real data analysis.}

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and generally improve with increased number of replicates per variety.}

### 2015-09-02 ebayes_rnaseq_heterosis.tex
fixed for Dan's comments

\begin{abstract}
An important type of heterosis, known as hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers are measuring transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. 
% This article has supplementary material online.
\end{abstract}

@@

Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means. We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH). Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, heterosis, and underdominance.  However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.  Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).  For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} to describe the situation where the hybrid mean is more extreme than the parental means, i.e. we say there is heterosis if and only if either HPH or LPH holds.

Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain poorly understood \citep{chen2013genomic}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents.  Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and is conceptually depicted in the right column of Figure 1b in  \cite{chen2013genomic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The results of read processing are summarized by a gene $\times$ sample matrix of counts. See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.

@@

We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}-\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.

@@

We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e. $\log(Y_{gvi}+1)-\gamma_{vi}$. Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.

@@

There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal). We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters and we discuss this in Section \ref{s:discussion}.

@@

As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions. Thus, there are no true values for the hyperparameters to compare against. But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values. 
\begin{table}
\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number. Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
\label{t:hyperparameter}
\begin{tabular}{l|r|rrr|}
& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{\# of replicates}\\
Parameter & value & 4 & 8 & 16 \\
\hline
% with (SD)
% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
% 3 decimal places
% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
$\eta_\phi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
\hline
$\sigma_\phi$ & 1.8\phantom{0} & 1.8\phantom{0}  & 1.8\phantom{0}  & 1.7\phantom{0} \\
$\sigma_\alpha$ & 0.36 & 0.46  & 0.40  & 0.36  \\
$\sigma_\delta$ & 0.15 & 0.39  & 0.32  & 0.27  \\
$\sigma_\psi$ & 0.74 & 0.40 & 0.49 & 0.54 \\
\hline
\end{tabular}
\end{table}
Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number. For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\phi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this. For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases. %We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates. For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.

@@

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes (Laplace) procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array}.
\right. 
\label{e:effect_size}
\end{equation}
This estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis. Through a simulation study, we demonstrated that this method outperformed alternative methods, and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the posterior via INLA. We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.

@@

In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e. the hyperparameters, will also improve detection of gene-expression heterosis. Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases. But typically the number of replicates per variety is quite small. In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient. There are a variety of alternative estimation approaches to explore, e.g. expectation-maximization algorithms or fully Bayesian approaches.

Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes involved in a genetic mechanism of heterosis.

>>>>

\begin{abstract}
An important type of heterosis, known as hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although hybrid vigor is extensively utilized in agriculture, its molecular basis  is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers are measuring transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. 
% This article has supplementary material online.
\end{abstract}

@@

Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means. We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH). Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, overdominance, and underdominance.  However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.  Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).  For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} (EH) to describe the situation where the hybrid mean is more extreme than the parental means, i.e. we say there is EH if and only if either HPH or LPH holds.

Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain poorly understood \citep{chen2013genomic}. One potential explanation, is EH of gene expression, i.e. enhanced (or suppressed) expression of one or more genes in hybrids compared to their inbred parents.  Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and EH of gene expression is conceptually depicted in the right column of Figure 1b in \cite{chen2013genomic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of EH. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where hybrid vigor is well established and identifies genes demonstrating EH of expression. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties and a cross between these two varieties called the hybrid. For each variety, replicate RNA samples are isolated and assessed for quality. Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The results of read processing are summarized by a gene $\times$ sample matrix of counts. See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.

To use \RNAseq{} counts to identify genes displaying EH of expression, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for EH.

@@

We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}--\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.

@@

We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one adjusted by the normalization factor, i.e. $\log(Y_{gvi}+1)-\gamma_{vi}$. Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.

@@

There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal). We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters, and we discuss this in Section \ref{s:discussion}.

@@

As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions. Thus, there are no true values for the hyperparameters to compare against. But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values. 
\begin{table}
\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number. Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
\label{t:hyperparameter}
\begin{tabular}{l|c|ccc|}
& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
Parameter & value & 4 & 8 & 16 \\
\hline
% with (SD)
% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
% 3 decimal places
% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
$\eta_\psi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
\hline
$\sigma_\phi$ & 1.8\phantom{00} & 1.8\phantom{00}  & 1.8\phantom{00}  & 1.7\phantom{00} \\
$\sigma_\alpha$ & 0.36\phantom{0} & 0.46\phantom{0}  & 0.40\phantom{0}  & 0.36\phantom{0}  \\
$\sigma_\delta$ & 0.15\phantom{0} & 0.39\phantom{0}  & 0.32\phantom{0}  & 0.27\phantom{0}  \\
$\sigma_\psi$ & 0.74\phantom{0} & 0.40\phantom{0} & 0.49\phantom{0} & 0.54\phantom{0} \\
\hline
\end{tabular}
\end{table}
Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number. For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this. For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases. %We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates. For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.

@@

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes (Laplace) procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute estimates of $\alpha_g$ and $\delta_g$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi_g$ estimates.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array}.
\right. 
\label{e:effect_size}
\end{equation}
This estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

@@

Geneticists speculate that gene expression EH is one possible explanation of hybrid vigor of traits, such as plant height or grain yield. Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting EH genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis. Through a simulation study, we demonstrated that this method outperformed alternative methods, and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the posterior via INLA. We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.

@@

In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e. the hyperparameters, will also improve detection of gene-expression heterosis. Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases. But typically the number of replicates per variety is quite small. In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient. There are a variety of alternative estimation approaches to explore, e.g. expectation-maximization algorithms or fully Bayesian approaches. Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes involved in a genetic mechanism of heterosis.
